A Deep Learning Approach for Gait Event Detection from a Single Shank-Worn IMU: Validation in Healthy and Neurological Cohorts

### 2022-04-13 paper/MDPI_template/article.tex
Sync for home office

% Abstract (Do not insert blank lines, i.e. \\) 
\abstract{A single paragraph of about 200 words maximum. For research articles, abstracts should give a pertinent overview of the work. We strongly encourage authors to use the following style of structured abstracts, but without headings: (1) Background: place the question addressed in a broad context and highlight the purpose of the study; (2) Methods: describe briefly the main methods or treatments applied; (3) Results: summarize the article's main findings; (4) Conclusions: indicate the main conclusions or interpretations. The abstract should be an objective representation of the article, it must not contain results which are not presented and substantiated in the main text and should not exaggerate the main conclusions.}

% Keywords
\keyword{keyword 1; keyword 2; keyword 3 (List three to ten pertinent keywords specific to the article; yet reasonably common within the subject discipline.)}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction \label{sec:Introduction}}
Gait deficits are common in older adults and possibly reflect the presence of an underlying neurodegenerative disease \cite{Snijders2007,Hodgins2008}. For example, conversion to Parkinson's Disease \cite{DelDin2019} or from mild cognitive impairment to Alzheimer's Disease \cite{Koenig2017,Bertoli2018} are linked with changes in spatiotemporal gait parameters. Similarly, temporal gait parameters are different for stroke patients \cite{SchroederVon1995,Mohan2021} and patients with multiple sclerosis \cite{Griskevicius2016,Flachenecker2019} when compared to healthy controls. To objectively quantify gait deficits, stride-specific parameters such as stride time or stride length are often used \cite{Hannink2016}, where the beginning and end of a stride are determined from two successive initial contacts (ICs) of the same foot \cite{Perry2010,Whittle2012}. The IC is when the foot contacts the ground and together with the instant at which the foot leaves the ground (final contact, FC), each stride can be divided in a stance and swing phase \cite{Rueterbories2010,Bruening2014}.  The events of IC and FC, also referred to as \emph{gait events}, are commonly determined using force or pressure measuring devices \cite{Bruening2014}, or passive, retro-reflective marker-based optoelectronic motion capture systems (henceforth referred to as the \emph{marker-based} system or method) \cite{Chiari2005,Topley2020}. These systems are relatively expensive, and restricted to usage in expertise laboratories \cite{Iosa2016,Jarchi2018}. As there is increasing evidence that gait measured in the lab does not reflect daily-life gait \cite{Hillel2019,Warmerdam2020,Atrsaei2021}, there is more and more interest in measurement systems that allow for continuous gait analysis in ambulatory settings. Therefore, the use of inertial measurement units (IMUs) is especially attractive, as these can be used to measure gait in ecologically valid environments, such as the home environment, thereby painting a more complete picture of health status \cite{DelDin2016,Shah2020} and providing clinical information that is complementary to standardized lab-based assessments \cite{Warmerdam2020,Fasano2020,Atrsaei2021,Corra2021}.

@@

In this study, we further these works by validating this deep learning approach for detecting gait events in a heterogeneous cohort of healthy and neurologically diseased adults considering a single IMU setup, worn on the lower leg.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Materials and Methods \label{sec:Materials_and_Methods}}
\subsection{Data Collection \label{subsec:Data_Collection}}
Gait analyses were performed in the Universit\"{a}tsklinikum Schleswig-Holstein (UKSH) campus Kiel, Germany. The study \cite{Warmerdam2021} was approved by the ethical committee of the medical faculty at the UKSH (no: D438/18). In total, data from 160 participants were included for the current analysis, including data from young adults (YA; age: 18 - 60 years), older adults (OA; age: $>$60 years), people with Parkinson's Disease (PD; according to the UK Brain Bank criteria \cite{Gibb1988}), people with a recent ($<$4 weeks) symptomatic stroke (stroke), people with multiple sclerosis (MS; according to the McDonalds criteria \cite{Thompson2018}), people with chronic low back pain (cLBP), and people with other diagnoses that were assumed not to affect mobility (Table \ref{tab:demographics_data}). Inclusion criteria were an age of 18 years or older, and the ability to walk independently without a walking aid. Participants were excluded from the study with a Montreal Cognitive Assessment \cite{Nasreddine2005} score $<$15 and other movement disorders that affected mobility, as noticed by the clinical assessor.

\begin{table}[H] 
	\caption{\label{tab:demographics_data}Demographics data of the study participants. Age, height, and weight are presented as mean (standard deviation).}
	\newcolumntype{C}{>{\centering\arraybackslash}X}
	\begin{tabularx}{\textwidth}{CCCCCC}
		\toprule
		\textbf{Group}	& \textbf{Gender}	& \textbf{Number of}	& \textbf{Age}	& \textbf{Height}	& \textbf{Weight}\\
		 & & \textbf{subjects} & years	& cm	& kg\\
		\midrule
		YA		& F		& 21	& 27 (7)	& 173 (5)	& 67 (8)\\
		 & M	& 22	& 30 (9)	& 185 (8) & 81 (13)\\ 
		\midrule
		OA		& F		& 11	& 70 (6)	& 167 (6)	& 72 (17)\\
		 & M	& 11	& 73 (6)	& 180 (6) & 83 (11)\\
		\midrule
		PD		& F		& 12	& 67 (6)	& 168 (6)	& 70 (14)\\
		 & M	& 20	& 62 (11)	& 178 (6) & 87 (13)\\
		\midrule
		MS		& F		& 12	& 37 (9)	& 174 (9)	& 75 (9)\\
		 & M	& 9	& 42 (15)	& 189 (8) & 96 (30)\\
		\midrule
		stroke	& F		& 4	& 66 (10)	& 160 (6)	& 65 (11)\\
		 & M	& 17	& 67 (17)	& 178 (7) & 84 (14)\\
		\midrule
		cLBP	& F		& 3	& 64 (10)	& 166 (5)	& 65 (5)\\
		 & M	& 7	& 63 (16)	& 178 (7) & 90 (16)\\ 
		\midrule
		other	& F		& 3	& 60 (13)	& 166 (3)	& 79 (15)\\
		 & M	& 8	& 68 (18)	& 182 (7) & 85 (13)\\ 
		\bottomrule
	\end{tabularx}
	\noindent{\footnotesize{\textbf{Group}: YA: younger adults, OA: older adults, PD: Parkinson's Disease, MS: multiple sclerosis, cLBP: chronic low back pain; \textbf{Gender}: F: female, M: male.}}
\end{table}

@@

Participants performed three walking trials consisting of walking 5 meter at either (1) preferred speed ("\emph{Please walk at your normal walking speed.}"), (2) slow speed ("\emph{Please walk half of your normal walking speed.}"), or (3) fast speed ("\emph{Please walk as fast as possible, without running or falling.}"). The 5 meter distance was marked with two cones on both ends, and participants were asked to start walking approximately two steps before the cones on one end, and stop walking approximately two steps after passing the cones on the other end.

@@

\subsection{Data Pre-processing \label{subsec:Data_Pre-processing}}
\subsubsection{Marker Data \label{subsubsec:Marker_Data}}
The reflective marker data were first used to determine start and end of each trial. The start of the trial was defined as the time for which the first toe marker (left or right) crossed the \emph{virtual} starting line. The end was defined as the time for which the last heel marker crossed the end line. For both marker and IMU systems, data were then cropped from start to the end of the trial.

Any remaining gaps in the marker data were filled by interpolation making use of inter-correlations between markers \cite{Federolf2013,Gloersen2016}. The data were then low-pass filtered using a double pass 6th order Butterworth filter with a cut-off frequency of 20 Hz \cite{Kormylo1974,Racz2021}. The filtered data were differentiated to get velocity signals, and timings of ICs and FCs were determined from local maxima and minima in the heel and toe vertical velocity signals \cite{Pijnappels2001,OConnor2007}. Like in \cite{Carcreff2018,Romijnders2021}, all identified ICs and FCs were manually checked, and corrected if necessary, using Qualisys Track Manager 2018.1 software (Qualisys AB, Göteborg, Sweden). The resulting annotated ICs and FCs were considered the \emph{true} events (also \emph{labels} or \emph{targets}), and were used as reference timings to derive stride-specific gait parameters.

@@

All participant data were split in three independent data sets, namely a training set, a validation set, and a test set. Each set contained data from approximately one-third of the participants. Participants were randomly assigned to one of the sets, stratified by both group (i.e., diagnosis) and gender (Table~\ref{tab:demographics_data}).

@@

\subsection{Analysis}
The predictions of the model on the test set data were compared with the labels from the test set. The model performance was evaluated for (1) overall detection performance, (2) time agreement between the predicted events and the (marker-based) annotated events, (3) agreement between subsequently derived gait parameters.
\subsubsection{Overall detection performance \label{subsubsec:overall_detection_performance}}
The overall detection performance quantified how many of the annotated evens were detected by the model (true positives, TP), how many of the annotated events were not detected (false negatives, FN), and how many event that were detected, were actually not annotated (false positives, FP). An annotated event was considered to be detected if any detected event was within 0.075 s. A detected event were defined as peaks in the predicted likelihoods with a minimum height of 0.4, i.e. $40\%$ likely that it corresponded to an event.

From these metrics, the recall (or sensitivity), precision and F$_{1}$ score were calculated as:
\begin{equation}
	\textrm{recall} = \frac{\textrm{TP}}{\textrm{TP} + \textrm{FN}} \nonumber
	\label{eqn:recall}
\end{equation}
\begin{equation}
	\textrm{precision} = \frac{\textrm{TP}}{\textrm{TP} + \textrm{FP}} \nonumber
	\label{eqn:precision}
\end{equation}
\begin{equation}
	\textrm{F$_{1}$ score} = 2 \cdot \frac{\textrm{recall} \cdot \textrm{precision}}{\textrm{recall} + \textrm{precision}}  \nonumber
	\label{eqn:f1_score}
\end{equation}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Results}

@@

Time agreement between the annotated and detected events was quantified for the TPs for each of the tracked points (Figure~\ref{fig:time_error_box_plots}). The median time error for each of the tracked points, and for both ICs and FCs was close to zero (Table~\ref{tab:time_errors}), and the largest deviation was -0.005 s, corresponding to one sample period. The IQR was at most 0.020 s, corresponding to 4 sample periods.
\begin{table}[H] 
	\caption{Time errors for the correctly detected gait events. \label{tab:time_errors}}
	\newcolumntype{C}{>{\centering\arraybackslash}X}
	\begin{tabularx}{\textwidth}{r|CC|CC}
		\toprule
		 & \multicolumn{2}{c}{\textbf{Initial contacts}} & \multicolumn{2}{c}{\textbf{Final contacts}}\\
		\textbf{Tracked point} & \textbf{median} & \textbf{IQR} & \textbf{median} & \textbf{IQR}\\
		 & s & s & s & s \\
		\midrule
		left ankle & 0.000 & 0.020 & 0.000 & 0.010\\
		right ankle & 0.000 & 0.020 & -0.005 & 0.015\\
		left shank & -0.005 & 0.020 & -0.005 & 0.020\\
		right shank & -0.003 & 0.020 & -0.005 & 0.020\\
		\bottomrule
	\end{tabularx}
	\noindent{\footnotesize{IQR: inter-quartile range.}}
\end{table}

@@

For all stride-specific gait parameters, and for all tracked points, the mean difference was close to zero, i.e., the maximum mean difference was 0.003 s, namely for the calculated swing time of the right ankle (Table~\ref{tab:stride_parameters}). Furthermore, for all gait parameters and for all tracked points the limits of agreement, based on a 95\% confidence interval were distributed around a zero mean difference with the overall limits of agreement at -0.049 s and 0.051 s (Figure~\ref{fig:gait_parameters_bland_altman_plots}).
\begin{table}[H] 
	\caption{Time agreement between the stride-specific parameters. \label{tab:stride_parameters}}
	\newcolumntype{C}{>{\centering\arraybackslash}X}
	\begin{tabularx}{\textwidth}{rrCC}
		\toprule
		\textbf{Tracked point} & \textbf{Parameters} & \textbf{Mean difference} & \textbf{Limits of Agreement}\\
		 & & ms & (ms, ms)\\
		\midrule
		left ankle & stride time & 0.001 & (-0.035, 0.036)\\
		 & stance time & 0.002 & (-0.039, 0.042)\\
		 & swing time & -0.001 & (-0.045, 0.043)\\
		\midrule
		right ankle & stride time & 0.000 & (-0.039, 0.040)\\
		 & stance time & -0.002 & (-0.048, 0.044)\\
		 & swing time & 0.003 & (-0.046, 0.051)\\
		\midrule
		left shank & stride time & 0.001 & (-0.039, 0.041)\\
		& stance time & 0.002 & (-0.043, 0.046)\\
		& swing time & -0.001 & (-0.049, 0.047)\\
		\midrule
		right shank & stride time & -0.000 & (-0.031, 0.031)\\
		& stance time & 0.002 & (-0.046, 0.049)\\
		& swing time & -0.002 & (-0.049, 0.046)\\
		\bottomrule
	\end{tabularx}
	\noindent{}
\end{table}

\begin{figure}[H]
	\centering
	\includegraphics[width=13.5cm]{fig/bland_altman_plots_stride_params.png}
	\caption{The agreement of extracted gait parameters between the sensor-based and marker-based methods.\label{fig:gait_parameters_bland_altman_plots}}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Discussion}
The current study aimed to validate a deep learning approach for detecting gait events from a single IMU worn on the lower leg. Data from a left and right ankle- and shank-worn IMU were used for training a deep learning model to detect gait events from walking trials performed by healthy younger and older adults, participants diagnosed with Parkinson's Disease, multiple sclerosis, or chronic low back pain, participants who had a recent symptomatic stroke, and participants diagnosed with other neurological diseases. Participants walked a 5 meter distance at three different self-selected walking speeds. The gait event timings that were predicted by the deep learning model were compared to a common reference method, i.e., optoelectronc motion capture, and clinically relevant stride-specific gait parameters were extracted.

@@

The main limitations of the current study were that only walking trials involving straight-line walking were considered, and the walking distance was relatively short. The observed gait patterns from these walking trials may therefore not fully representative of gait patterns observed in daily-life \cite{Hillel2019,Warmerdam2020,Atrsaei2021}. However, as the deep learning-based approach does not rely on fixed thresholds or assumptions regarding to which sensor axis is used, it is theoretically transferable and scalable to other conditions if input data and corresponding labels can be provided. For example, in \cite{Gadaleta2019} a similar deep learning model was tested for a setup with three IMUs and for a setup with a single IMU, and observed differences in time errors compared to a pressure sensing walkway were negligible.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Conclusions}
In this study we have validated a deep learning-based approach to detect gait events and subsequently extract clinically relevant stride parameters from a single inertial measurement unit worn either laterally above the ankle joint or proximally below the knee joint. Performance analysis showed an excellent detection rate, and low time errors in both event detection and stride parameter calculation for different walking speeds and across both healthy and neurological cohorts. Our next step is to validate these methods with real-life walking sequences.

@@

\dataavailability{Data from the first 10 participant are available online at \url{https://github.com/neurogeriatricskiel/Validation-dataset}. Additionally, we are preparing the open-source release of all data from the healthy younger and older adults. Data from ptient groups }

\acknowledgments{The authors sincerely thank all people that were involved in the data collection, from study participants to students who aided and assisted during measurements and subsequent marker labeling. The authors appreciate the off- and on-topic discussions with Julius Welzel, and his positive contributions to the organization of the research data. The authors give a thumbs up to Johannes Hoffmann for his tips and tricks regarding the \LaTeX typesetting.}

>>>>

% Abstract (Do not insert blank lines, i.e. \\) 
\abstract{Many algorithms use 3D accelerometer and/or gyroscope data from IMU sensors to detect gait events (i.e., initial and final contact). However, these algorithms often require knowledge about sensor orientation and use empirically derived thresholds. As alignment cannot always be controlled for in ambulatory assessments, methods are needed that require little knowledge on sensor location and orientation, e.g., a convolutional neural network-based deep learning model. Therefore, 160 participants from healthy and neurologically diseased cohorts walked 5 meter distances at slow, preferred, and fast walking speed, while data were collected from IMUs on the left and right ankle and shank. Gait events were detected and stride parameters were extracted using a deep learning model and an optoelectronic motion capture (OMC) system for reference. Results showed a high detection rate for both initial and final contact detection across the IMUs (recall~$\ge 92\%$, precision~$\ge 97\%$). Time agreement was excellent, as witnessed median time error (0.005 s) and corresponding inter-quartile range (0.020~s). The extracted stride-specific parameters were in good agreement with parameters derived from the OMC system (maximum mean difference 0.003~s and corresponding maximum limits of agreement (-0.049, 0.051)~s for $95\%$ confidence level). Thus, the deep learning approach was considered a valid approach for detecting gait events and extracting stride-specific parameters with little knowledge on exact IMU location and orientation in conditions with and without walking pathologies due to neurological diseases.}

% Keywords
\keyword{gait; inertial measurement units; deep learning}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction \label{sec:Introduction}}
Gait deficits are common in older adults and possibly reflect the presence of an underlying neurodegenerative disease \cite{Snijders2007,Hodgins2008}. For example, conversion to Parkinson's Disease \cite{DelDin2019} or from mild cognitive impairment to Alzheimer's Disease \cite{Koenig2017,Bertoli2018} are linked with changes in spatiotemporal gait parameters. Similarly, temporal gait parameters are different for stroke patients \cite{SchroederVon1995,Mohan2021} and patients with multiple sclerosis \cite{Griskevicius2016,Flachenecker2019} when compared to healthy controls. To objectively quantify gait deficits, stride-specific parameters such as stride time or stride length are often used \cite{Hannink2016}. The beginning and end of a stride are determined from two successive initial contacts (ICs) of the same foot \cite{Perry2010,Whittle2012}. The IC is when the foot contacts the ground, and together with the instant at which the foot leaves the ground (final contact, FC), each stride can be divided in a stance and swing phase \cite{Rueterbories2010,Bruening2014}.  The events of IC and FC, also referred to as \emph{gait events}, are commonly determined using force or pressure measuring devices \cite{Bruening2014}, or marker-based optoelectronic motion capture systems (henceforth referred to as the \emph{marker-based} system or method) \cite{Chiari2005,Topley2020}. These systems are relatively expensive, and restricted to usage in expertise laboratories \cite{Iosa2016,Jarchi2018}. As there is increasing evidence that gait measured in the lab does not reflect daily-life gait \cite{Hillel2019,Warmerdam2020,Atrsaei2021}, there is increasingly more interest in measurement systems that allow for continuous gait analysis in ambulatory settings. Therefore, the use of inertial measurement units (IMUs) is especially attractive, as these can be used to measure gait in ecologically valid environments, such as the home environment, thereby painting a more complete picture of health status \cite{DelDin2016,Shah2020} and providing clinical information that is complementary to standardized lab-based assessments \cite{Warmerdam2020,Fasano2020,Atrsaei2021,Corra2021}.

@@

The aim of this study is to extend these works by validating this deep learning approach for detecting gait events in a heterogeneous cohort of healthy and neurologically diseased adults considering a single IMU setup, worn on the lower leg.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Materials and Methods \label{sec:Materials_and_Methods}}
\subsection{Data Collection \label{subsec:Data_Collection}}
Gait analyses were performed in the Universit\"{a}tsklinikum Schleswig-Holstein (UKSH) campus Kiel, Germany. The study \cite{Warmerdam2021} was approved by the ethical committee of the medical faculty at the UKSH (no: D438/18). In total, data from 160 participants were included for the current analysis, including data from young adults (YA; age: 18 - 60 years), older adults (OA; age: $>$60 years), people with Parkinson's Disease (PD; according to the UK Brain Bank criteria \cite{Gibb1988}), people with a recent ($<$4 weeks) symptomatic stroke (stroke), people with multiple sclerosis (MS; according to the McDonalds criteria \cite{Thompson2018}), people with chronic low back pain (cLBP), and people with other diagnoses that affected mobility (Table \ref{tab:demographics_data}). Inclusion criteria were an age of 18 years or older, and the ability to walk independently without a walking aid. Participants were excluded from the study with a Montreal Cognitive Assessment \cite{Nasreddine2005} score $<$15 and other movement disorders that affected mobility, as noticed by the clinical assessor.

\begin{table}[H] 
	\caption{\label{tab:demographics_data}Demographics data of the study participants. Age, height, and weight are presented as mean (standard deviation).}
	\newcolumntype{C}{>{\centering\arraybackslash}X}
	\begin{tabularx}{\textwidth}{CCCCCC}
		\toprule
		\textbf{Group}	& \textbf{Gender}	& \textbf{Number of}	& \textbf{Age}	& \textbf{Height}	& \textbf{Weight}\\
		 & & \textbf{participants} & years	& cm	& kg\\
		\midrule
		YA		& F		& 21	& 27 (7)	& 173 (5)	& 67 (8)\\
		 & M	& 22	& 30 (9)	& 185 (8) & 81 (13)\\ 
		\midrule
		OA		& F		& 11	& 70 (6)	& 167 (6)	& 72 (17)\\
		 & M	& 11	& 73 (6)	& 180 (6) & 83 (11)\\
		\midrule
		PD		& F		& 12	& 67 (6)	& 168 (6)	& 70 (14)\\
		 & M	& 20	& 62 (11)	& 178 (6) & 87 (13)\\
		\midrule
		MS		& F		& 12	& 37 (9)	& 174 (9)	& 75 (9)\\
		 & M	& 9	& 42 (15)	& 189 (8) & 96 (30)\\
		\midrule
		stroke	& F		& 4	& 66 (10)	& 160 (6)	& 65 (11)\\
		 & M	& 17	& 67 (17)	& 178 (7) & 84 (14)\\
		\midrule
		cLBP	& F		& 3	& 64 (10)	& 166 (5)	& 65 (5)\\
		 & M	& 7	& 63 (16)	& 178 (7) & 90 (16)\\ 
		\midrule
		other	& F		& 3	& 60 (13)	& 166 (3)	& 79 (15)\\
		 & M	& 8	& 68 (18)	& 182 (7) & 85 (13)\\ 
		\bottomrule
	\end{tabularx}
	\noindent{\footnotesize{\textbf{Group}: YA: younger adults, OA: older adults, PD: Parkinson's Disease, MS: multiple sclerosis, cLBP: chronic low back pain; \textbf{Gender}: F: female, M: male.}}
\end{table}

@@

Participants performed three walking trials consisting of walking 5 meter at either (1) preferred speed ("\emph{Please walk at your normal walking speed.}"), (2) slow speed ("\emph{Please walk at half of your normal walking speed.}"), or (3) fast speed ("\emph{Please walk as fast as possible, without running or falling.}"). The 5 meter distance was marked with two cones on both ends, and participants were asked to start walking approximately two steps before the cones on one end, and stop walking approximately two steps after passing the cones on the other end.

@@

\subsection{Data Pre-processing \label{subsec:Data_Pre-processing}}
\subsubsection{Marker Data \label{subsubsec:Marker_Data}}
First, data from both marker and IMU systems were cropped so that only data from within the 5 meter distance was considered. Any gaps in the marker data were then filled by interpolation making use of inter-correlations between markers \cite{Federolf2013,Gloersen2016}. The data were then low-pass filtered using a double pass 6th order Butterworth filter with a cut-off frequency of 20 Hz \cite{Kormylo1974,Racz2021}. The filtered data were differentiated to get velocity signals, and timings of ICs and FCs were determined from local maxima and minima in the heel and toe vertical velocity signals \cite{Pijnappels2001,OConnor2007}. Like in \cite{Carcreff2018,Romijnders2021}, all identified ICs and FCs were manually checked using Qualisys Track Manager 2018.1 software (Qualisys AB, Göteborg, Sweden), and corrected if necessary. The resulting annotated ICs and FCs were considered the \emph{true} events (also \emph{labels} or \emph{targets}), and were used as reference timings to derive stride-specific gait parameters.

@@

All participant data were split in three independent data sets, namely a training set, a validation set, and a test set. Each set contained data from approximately one-third of the participants. Participants were randomly assigned to one of the sets, stratified by both group (i.e., diagnosis) and gender (Table~\ref{tab:demographics_data}). The training and validation set were used to train an optimal deep learning model, whereas the test set was used to evaluate the model performance on \emph{unseen} data (that is, the test set data were not used to train or find the model architecture). Any results on the model performance therefore, were only based on the test set (Section~\ref{sec:Results}).

@@

\subsection{Analysis}
The predictions of the model on the test set data were compared with the labels from the test set. The model performance was evaluated for (1) overall detection performance, (2) time agreement between the predicted events and the (marker-based) annotated events, (3) agreement between subsequently derived gait parameters.
\subsubsection{Overall detection performance \label{subsubsec:overall_detection_performance}}
The overall detection performance quantified how many of the annotated events were detected by the model (true positives, TP), how many of the annotated events were not detected (false negatives, FN), and how many event that were detected, were actually not annotated (false positives, FP). An annotated event was considered to be detected if any detected event was within 0.075 s. A detected event was defined as peaks in the predicted likelihoods with a minimum height of 0.4, i.e. $40\%$ likely that it corresponded to an event.

From these metrics, the recall (or sensitivity), precision and F$_{1}$ score were calculated as:
\begin{equation}
	\textrm{recall} = \frac{\textrm{TP}}{\textrm{TP} + \textrm{FN}} 
	\label{eqn:recall}
\end{equation}
\begin{equation}
	\textrm{precision} = \frac{\textrm{TP}}{\textrm{TP} + \textrm{FP}} 
	\label{eqn:precision}
\end{equation}
\begin{equation}
	\textrm{F$_{1}$ score} = 2 \cdot \frac{\textrm{recall} \times \textrm{precision}}{\textrm{recall} + \textrm{precision}}  
	\label{eqn:f1_score}
\end{equation}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Results}
\label{sec:Results}

@@

Time agreement between the annotated and detected events was quantified for the TPs for each of the tracked points (Figure~\ref{fig:time_error_box_plots}). The median time error for each of the tracked points, and for both ICs and FCs was close to zero (Table~\ref{tab:time_errors}), and the largest deviation was -0.005 s, corresponding to one sample period (at the sampling frequency of 200 Hz). The IQR was at most 0.020 s, corresponding to 4 sample periods.
\begin{table}[H] 
	\caption{Time errors for the correctly detected gait events. \label{tab:time_errors}}
	\newcolumntype{C}{>{\centering\arraybackslash}X}
	\begin{tabularx}{\textwidth}{r|CC|CC}
		\toprule
		 & \multicolumn{2}{c}{\textbf{Initial contacts}} & \multicolumn{2}{c}{\textbf{Final contacts}}\\
		\textbf{Tracked point} & \textbf{median} & \textbf{IQR} & \textbf{median} & \textbf{IQR}\\
		 & s & s & s & s \\
		\midrule
		left ankle & 0.000 & 0.020 & 0.000 & 0.010\\
		right ankle & 0.000 & 0.020 & -0.005 & 0.015\\
		left shank & -0.005 & 0.020 & -0.005 & 0.020\\
		right shank & -0.003 & 0.020 & -0.005 & 0.020\\
		\bottomrule
	\end{tabularx}
	\noindent{\footnotesize{IQR: inter-quartile range.}}
\end{table}

@@

For all stride-specific gait parameters, and for all tracked points, the mean difference was close to zero, i.e., the maximum mean difference was 0.003 s, namely for the calculated swing time of the right ankle (Table~\ref{tab:stride_parameters}). Furthermore, for all gait parameters and for all tracked points the limits of agreement, based on a 95\% confidence interval were distributed around a zero mean difference with the overall limits of agreement at -0.049 s and 0.051 s (Figure~\ref{fig:gait_parameters_bland_altman_plots}).
\begin{table}[H] 
	\caption{Time agreement between the stride-specific parameters. \label{tab:stride_parameters}}
	\newcolumntype{C}{>{\centering\arraybackslash}X}
	\begin{tabularx}{\textwidth}{rrCC}
		\toprule
		\textbf{Tracked point} & \textbf{Parameters} & \textbf{Mean difference} & \textbf{Limits of Agreement}\\
		 & & s & s\\
		\midrule
		left ankle & stride time & 0.001 & (-0.035, 0.036)\\
		 & stance time & 0.002 & (-0.039, 0.042)\\
		 & swing time & -0.001 & (-0.045, 0.043)\\
		\midrule
		right ankle & stride time & 0.000 & (-0.039, 0.040)\\
		 & stance time & -0.002 & (-0.048, 0.044)\\
		 & swing time & 0.003 & (-0.046, 0.051)\\
		\midrule
		left shank & stride time & 0.001 & (-0.039, 0.041)\\
		& stance time & 0.002 & (-0.043, 0.046)\\
		& swing time & -0.001 & (-0.049, 0.047)\\
		\midrule
		right shank & stride time & -0.000 & (-0.031, 0.031)\\
		& stance time & 0.002 & (-0.046, 0.049)\\
		& swing time & -0.002 & (-0.049, 0.046)\\
		\bottomrule
	\end{tabularx}
	\noindent{}
\end{table}

\begin{figure}[H]
	\begin{adjustwidth}{-\extralength}{0cm}
		\centering
		\includegraphics[width=16.5cm]{fig/bland_altman_plots_stride_params.png}
	\end{adjustwidth}
	\caption{The agreement of extracted gait parameters between the sensor-based and marker-based methods.\label{fig:gait_parameters_bland_altman_plots}}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Discussion}
The current study aimed to validate a deep learning approach for detecting gait events from a single IMU worn on the lower leg. Data from a left and right ankle- and shank-worn IMU were used for training a deep learning model to detect gait events from walking trials performed by healthy YA, OA, participants diagnosed with PD, MS, or cLBP, participants who had a recent symptomatic stroke, and participants diagnosed with other neurological diseases. Participants walked a 5 meter distance at three different self-selected walking speeds. The gait event timings that were predicted by the deep learning model were compared to a common reference method, i.e., optoelectronic motion capture, and clinically relevant stride-specific gait parameters were extracted.

@@

The main limitations of the current study were that only walking trials involving straight-line walking were considered, and the walking distance was relatively short. The observed gait patterns from these walking trials may therefore not be fully representative of gait patterns observed in daily-life \cite{Hillel2019,Warmerdam2020,Atrsaei2021}. However, as the deep learning-based approach does not rely on fixed thresholds or assumptions regarding to which sensor axis is used, it is theoretically transferable and scalable to other conditions if input data and corresponding labels can be provided. For example, in \cite{Gadaleta2019} a similar deep learning model was tested for a setup with three IMUs and for a setup with a single IMU, and observed differences in time errors compared to a pressure sensing walkway were negligible.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Conclusions}
In this study we have validated a deep learning-based approach to detect gait events and subsequently extract clinically relevant stride parameters from a single inertial measurement unit worn either laterally above the ankle joint or proximal below the knee joint. Performance analysis showed an excellent detection rate, and low time errors in both event detection and stride parameter calculation for different walking speeds and across both healthy and neurological cohorts. Our next step is to validate these methods with real-life walking sequences.

@@

\dataavailability{Data from the first 10 participant are available online at \url{https://github.com/neurogeriatricskiel/Validation-dataset}. Additionally, we are preparing the open-source release of all data from the healthy younger and older adults. Data from patient groups can be shared upon reasonable request. The full algorithm pipeline is going to be made publicly available online at \url{https://github.com/rmndrs89/my-gait-events-tcn}.}

\acknowledgments{The authors sincerely thank all people that were involved in the data collection, from study participants to students who aided and assisted during measurements and subsequent marker labeling. The authors thank Julius Welzel for his contributions to the organization of the research data. The authors thank Johannes Hoffmann for his tips regarding \LaTeX\ typesetting.}
